---
title: "Choosing a thinning method"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing a thinning method}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(thinr)
```

Thinning reduces a binary shape to a one-pixel-wide centerline that preserves topology. Several algorithms exist; they differ in how aggressively they erode pixels, how well they preserve corners, and how fast they run. This vignette gives a short guide to choosing among the algorithms `thinr` provides.

## What's implemented

| Method        | Status (v0.2)   | Reference |
|---------------|-----------------|-----------|
| `zhang_suen`  | Full            | Zhang & Suen (1984) |
| `guo_hall`    | Full            | Guo & Hall (1989) |
| `lee`         | Full (2-D)      | Lee, Kashyap & Chu (1994) |
| `k3m`         | Full            | Saeed et al. (2010) |

`zhang_suen` is the default and matches `EBImage::thinImage()` behavior. The `thinImage()` wrapper is provided as a drop-in replacement.

`lee` is a 2-D adaptation of Lee's original 3-D algorithm. The 3-D case (volumetric thinning) is not implemented in this release.

The K3M lookup tables in this implementation are reconstructed from the algorithm's published description; they produce topology-preserving, one-pixel-wide skeletons on the test corpus. Verification against the paper's exact tables is welcomed.

## A quick visual comparison

```{r}
# A 'V' shape — exercises diagonal preservation
make_v <- function() {
  m <- matrix(0L, 11, 11)
  for (k in seq(0, 5)) {
    m[2 + k, 2 + k] <- 1L
    m[2 + k, 10 - k] <- 1L
    m[3 + k, 2 + k] <- 1L
    m[3 + k, 10 - k] <- 1L
  }
  m
}
v <- make_v()
v
```

```{r}
thin(v, method = "zhang_suen")
```

```{r}
thin(v, method = "guo_hall")
```

```{r}
thin(v, method = "lee")
```

```{r}
thin(v, method = "k3m")
```

All four algorithms produce one-pixel-wide skeletons that follow the diagonals. Differences show up at corners and small protrusions: Guo-Hall and K3M tend to preserve corners marginally better than Zhang-Suen; Lee tends toward thinner skeletons because its four directional sub-iterations process boundaries more aggressively.

## When to use which

- **`zhang_suen`** — the default. Most predictable behavior. Use for general purpose thinning and for compatibility with code that previously used `EBImage::thinImage()`.
- **`guo_hall`** — try this if your skeletons have lots of diagonal features and Zhang-Suen is breaking them at corners. Slightly different topology preservation behavior.
- **`lee`** — when you want directional processing (four sub-iterations per pass, one per cardinal direction). Sometimes produces cleaner skeletons on asymmetric inputs than Zhang-Suen's two-subiteration approach.
- **`k3m`** — strongest corner preservation in published comparative studies, at the cost of being slower (six phases per outer iteration vs. two for Zhang-Suen).

## Inputs and outputs

`thin()` accepts logical, integer, and numeric matrices. Non-zero values are foreground. The return matrix matches the storage mode of the input — logical in, logical out; integer in, integer out; numeric in, numeric out.

Higher-dimensional arrays (e.g. 3-D volumes) are not yet supported. The Lee algorithm in v0.2 will add 3-D support.

## Performance

A simple `bench::mark()` comparison on a moderate image (illustrative — see `tests/testthat/test-thin.R` for the test corpus that is the basis for the benchmark report committed alongside this release):

```{r eval = FALSE}
library(bench)
m <- matrix(0L, 200, 200)
m[50:150, 50:150] <- 1L  # solid square

bench::mark(
  zs = thin(m, method = "zhang_suen"),
  gh = thin(m, method = "guo_hall"),
  iterations = 5,
  check = FALSE
)
```

All four algorithms are O(iterations × pixels). Constant factors increase roughly: `zhang_suen` < `guo_hall` < `lee` < `k3m`. For 200×200 images all four finish in a few milliseconds on modern hardware.

## References

- Zhang, T. Y. & Suen, C. Y. (1984). A fast parallel algorithm for thinning digital patterns. *Communications of the ACM*, 27(3), 236–239.
- Guo, Z. & Hall, R. W. (1989). Parallel thinning with two-subiteration algorithms. *Communications of the ACM*, 32(3), 359–373.
- Lee, T.-C., Kashyap, R. L., & Chu, C.-N. (1994). Building skeleton models via 3-D medial surface axis thinning algorithms. *CVGIP: Graphical Models and Image Processing*, 56(6), 462–478.
- Saeed, K., Tabędzki, M., Rybnik, M., & Adamski, M. (2010). K3M: A universal algorithm for image skeletonization and a review of thinning techniques. *International Journal of Applied Mathematics and Computer Science*, 20(2), 317–335.
